---
title: "Gaming vs. Academic Performance"
subtitle: "INFO 432 — Group 5 Final Report"
author: "Jeffrey Cheung, Drew Jones, Danny Eapen, Joel Thomas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    number_sections: true
    code_folding: hide
    fig_width: 8
    fig_height: 5
    self_contained: true
---

```{r setup, include = FALSE}
# Every path below is relative to the project root, not to report/, so that the
# same code works whether it is knit from RStudio or run through Rscript.
knitr::opts_knit$set(root.dir = normalizePath(".."))

knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  dpi     = 110,
  # html_document renders figures at 2x by default. With 8,000-point scatterplots
  # that pushes the self-contained HTML past 9 MB, so render at 1x instead.
  fig.retina = 1
)

library(tidyverse)
library(corrplot)
library(rpart)
library(rpart.plot)
library(broom)

# The full one-way ANOVA block in section 4.4 fits thousands of factor levels and
# takes a couple of minutes. Set to FALSE for a fast knit; the method-comparison
# table is then skipped.
RUN_ANOVA <- TRUE

# Shared helper so every table in the report is formatted the same way.
tbl <- function(x, caption = NULL, digits = 3) {
  knitr::kable(x, caption = caption, digits = digits, format = "html",
               table.attr = 'class="table table-striped table-condensed"')
}

set.seed(432)
```

```{r output-dir, include = FALSE}
dir.create("output", showWarnings = FALSE)
```

---

# Introduction

## Research questions

Video games are the default leisure activity for a large share of college students, and
the folk claim that gaming "ruins your grades" is repeated far more often than it is
measured. This report takes a dataset of 8,000 students and asks three questions:

1. **Which lifestyle factors actually relate to grades?** Gaming hours, study hours,
   sleep hours, class attendance, and a self-reported gaming addiction score are the five
   candidates. We test each one for a statistically significant relationship with grades.
2. **Where are the thresholds?** A significant correlation says a relationship exists but
   not where it starts to matter. We look for the specific values — hours of gaming per
   day, hours of study — at which average performance visibly steps down.
3. **How well can grades be predicted?** We build and compare regression models, and
   score them on students the model never saw, so the reported accuracy is honest.

## The dataset

The data comes from the Kaggle dataset
[Gaming vs Academic Performance](https://www.kaggle.com/datasets/nalisha/gaming-vs-academic-performance):
8,000 student records with 14 columns covering demographics, gaming habits, study
habits, sleep, attendance, stress, and a final grade.

```{r load-raw}
df_raw <- read_csv("data/Gaming_Academic_Performance.csv", show_col_types = FALSE)

tibble(
  Column = names(df_raw),
  Type   = map_chr(df_raw, ~ class(.x)[1]),
  Example = map_chr(df_raw, ~ as.character(.x[1]))
) %>%
  tbl(caption = "Raw dataset structure (8,000 rows x 14 columns)")
```

A caveat we state up front and return to in the limitations: this dataset is almost
certainly **simulated rather than collected**. The relationships in it are far cleaner
than real survey data ever is. Everything below is a correct application of the method to
the data as given; none of it is a real-world claim about students.

---

# Data cleaning

*Source script: [`scripts/cleaning.R`](../scripts/cleaning.R)*

## Missing values

```{r missing}
missing <- colSums(is.na(df_raw))

tibble(Column = names(missing), `Missing values` = as.integer(missing)) %>%
  tbl(caption = "Missing value count by column")
```

## The grades problem

The grade column runs past 100, which is not possible on a normal percentage scale.

```{r grades-issue}
over_100 <- df_raw %>% filter(grades > 100)

tibble(
  Statistic = c("Minimum", "Median", "Mean", "Maximum",
                "Records over 100", "Percent of sample"),
  Value = c(
    round(min(df_raw$grades), 2),
    round(median(df_raw$grades), 2),
    round(mean(df_raw$grades), 2),
    round(max(df_raw$grades), 2),
    nrow(over_100),
    round(100 * nrow(over_100) / nrow(df_raw), 2)
  )
) %>%
  tbl(caption = "Grade distribution before cleaning")
```

Before deciding what to do with these 134 records, we checked whether they look like data
entry errors or like genuine high performers:

```{r over-100-profile}
bind_rows(
  over_100 %>%
    summarise(across(c(gaming_hours, study_hours, sleep_hours, attendance,
                       addiction_score), mean)) %>%
    mutate(Group = "Grades over 100", .before = 1),
  df_raw %>%
    filter(grades <= 100) %>%
    summarise(across(c(gaming_hours, study_hours, sleep_hours, attendance,
                       addiction_score), mean)) %>%
    mutate(Group = "Everyone else", .before = 1)
) %>%
  tbl(caption = "Profile of the over-100 records vs. the rest of the sample", digits = 2)
```

They study more, game less, attend more, and score lower on addiction — exactly the
profile of the strongest students. So these are informative cases, not noise. **We capped
them at 100 rather than dropping them**, which keeps their predictor values in the sample
at the cost of censoring their outcome. That censoring is carried forward as a known
limitation in the modeling section.

```{r clean}
df <- df_raw %>%
  mutate(grades = pmin(grades, 100)) %>%
  mutate(
    gender        = as.factor(gender),
    gaming_genre  = as.factor(gaming_genre),
    stress_level  = factor(stress_level, levels = c("Low", "Medium", "High"),
                           ordered = TRUE)
  )

write_csv(df, "data/cleaned_data.csv")
```

## Post-cleaning distribution

```{r grade-hist, fig.height = 5}
p_hist <- ggplot(df, aes(x = grades)) +
  geom_histogram(binwidth = 5, color = "black", fill = "steelblue") +
  stat_bin(binwidth = 5, geom = "text",
           aes(label = ifelse(after_stat(count) > 400, after_stat(count), "")),
           vjust = -0.5, size = 3.5) +
  labs(title = "Distribution of grades after cleaning",
       subtitle = "Counts labelled where a bin holds more than 400 students",
       x = "Grade", y = "Number of students") +
  theme_minimal()

ggsave("output/grade_distribution_graph.png", p_hist, width = 8, height = 5, dpi = 110)
p_hist
```

The distribution is roughly normal with a visible spike at the top — that spike is the
capping we just applied.

---

# Exploratory data analysis

*Source script: [`scripts/eda.R`](../scripts/eda.R)*

## Correlation matrix

```{r corr-matrix, fig.height = 6.5, fig.width = 7}
numeric_vars <- df %>% select(where(is.numeric)) %>% select(-student_id)
corr_matrix  <- cor(numeric_vars, use = "complete.obs")

png("output/correlation_matrix.png", width = 1400, height = 1300, res = 170)
corrplot(corr_matrix, method = "color", type = "upper", tl.cex = 0.7,
         addCoef.col = "black", number.cex = 0.55)
invisible(dev.off())

corrplot(corr_matrix, method = "color", type = "upper", tl.cex = 0.8,
         addCoef.col = "black", number.cex = 0.6)
```

Two things stand out immediately and shape the rest of the analysis:

```{r corr-highlights}
tibble(
  Pair = c("study_hours ~ grades", "gaming_hours ~ grades",
           "addiction_score ~ grades", "attendance ~ grades",
           "sleep_hours ~ grades", "addiction_score ~ gaming_hours"),
  Correlation = c(
    cor(df$study_hours, df$grades), cor(df$gaming_hours, df$grades),
    cor(df$addiction_score, df$grades), cor(df$attendance, df$grades),
    cor(df$sleep_hours, df$grades), cor(df$addiction_score, df$gaming_hours)
  )
) %>%
  tbl(caption = "Correlations worth naming")
```

First, study hours and gaming hours are the two dominant relationships and they point in
opposite directions. Second, **addiction score correlates 0.91 with gaming hours** — those
two columns are close to the same measurement, which becomes the central issue in the
regression section.

## Grades vs. each factor

```{r scatterplots, fig.height = 4.5}
plot_vars <- c("gaming_hours", "study_hours", "sleep_hours", "attendance",
               "addiction_score")

for (v in plot_vars) {
  p <- ggplot(df, aes(x = .data[[v]], y = grades)) +
    geom_point(alpha = 0.25, size = 0.8) +
    geom_smooth(method = "loess", color = "blue", se = TRUE) +
    labs(title = paste("Grades vs", v), x = v, y = "Grade") +
    theme_minimal()

  # Explicit size, otherwise ggsave inherits the device dimensions and writes
  # multi-megabyte files.
  ggsave(paste0("output/scatter_", v, ".png"), plot = p,
         width = 8, height = 5, dpi = 100)
  print(p)
}
```

The gaming and addiction curves **bend** rather than run straight — the penalty per extra
hour is not constant. That is the motivation for the squared terms in the modeling
section.

## Grades by categorical variables

```{r boxplots, fig.height = 4.5}
ggplot(df, aes(x = stress_level, y = grades, fill = stress_level)) +
  geom_boxplot(alpha = 0.8) +
  scale_fill_brewer(palette = "YlOrRd") +
  labs(title = "Grades by stress level", x = "Stress level", y = "Grade") +
  theme_minimal() +
  theme(legend.position = "none")

ggplot(df, aes(x = gaming_genre, y = grades, fill = gaming_genre)) +
  geom_boxplot(alpha = 0.8) +
  labs(title = "Grades by gaming genre", x = "Gaming genre", y = "Grade") +
  theme_minimal() +
  theme(legend.position = "none")
```

Neither categorical variable separates students the way the continuous factors do, so the
analysis from here on focuses on the five numeric factors.

---

# Statistical testing

*Source script: [`scripts/stat_testing.R`](../scripts/stat_testing.R)*

**Setup.** We use $\alpha = 0.05$. For each factor the null hypothesis is that it has no
relationship with grades; the alternative is that it does.

```{r vars}
vars <- c("gaming_hours", "study_hours", "sleep_hours", "attendance",
          "addiction_score")
```

## A note on method: why we dropped ANOVA

Our first version of this analysis used one-way ANOVA with `factor(gaming_hours)`,
`factor(attendance)`, and so on. That was the wrong tool, and it gave us wrong answers.
ANOVA compares a handful of groups, but these factors are continuous measurements, so
`factor()` creates one group per distinct value:

```{r cell-sizes}
cell_sizes <- tibble(
  variable     = vars,
  n_levels     = map_int(vars, ~ length(unique(df[[.x]]))),
  obs_per_cell = round(map_dbl(vars, ~ nrow(df) / length(unique(df[[.x]]))), 2)
)

tbl(cell_sizes, caption = "Distinct values per factor — why the ANOVA breaks down",
    digits = 2)
```

Attendance has 3,472 distinct values across 8,000 students — roughly **two students per
group**. With cells that small there is almost nothing left to average within a group,
which is exactly what the F test needs, and the test reported attendance as *not*
significant when it actually is. The correlation tests and the regression below are the
results we use; the ANOVA is kept in section 4.4 purely to document the comparison.

## Test 1 — correlation of each factor with grades

This is the direct replacement for the one-way ANOVA. It asks the same question, "does
this factor relate to grades", but it uses the fact that the values are *ordered* instead
of throwing that information away.

```{r cor-tests}
cor_results <- map_dfr(vars, function(v) {
  ct <- cor.test(df[[v]], df$grades)
  m  <- lm(df$grades ~ df[[v]])
  tibble(
    variable   = v,
    r          = round(unname(ct$estimate), 3),
    p_value    = ct$p.value,
    slope      = round(unname(coef(m)[2]), 3),
    r_squared  = round(summary(m)$r.squared, 4),
    conclusion = ifelse(ct$p.value < 0.05, "Reject H0", "Fail to reject")
  )
})

write_csv(cor_results, "output/correlation_results.csv")

cor_results %>%
  mutate(p_value = format.pval(p_value, digits = 3, eps = 1e-300)) %>%
  tbl(caption = "Each factor tested against grades on its own")
```

**Every factor, including attendance, is significant on its own.** The `slope` column is
the part worth quoting in a summary: it converts a p-value into plain language — one extra
hour of study is worth about `r round(cor_results$slope[cor_results$variable == "study_hours"], 1)`
grade points, one extra hour of gaming about
`r round(cor_results$slope[cor_results$variable == "gaming_hours"], 1)`.

## Test 2 — all five factors in one regression

The correlations above look at one factor at a time, so a factor can look important just
by riding along with another one. Putting them in the same model shows what each factor is
worth *once the others are already accounted for*.

```{r full-model}
full_model <- lm(
  grades ~ gaming_hours + study_hours + sleep_hours + attendance + addiction_score,
  data = df
)

reg_results <- tidy(full_model) %>%
  mutate(conclusion = ifelse(p.value < 0.05, "Reject H0", "Fail to reject"))

write_csv(reg_results, "output/regression_results.csv")

reg_results %>%
  mutate(p.value = format.pval(p.value, digits = 3, eps = 1e-300)) %>%
  tbl(caption = sprintf("Multiple regression, all five factors (adjusted R-squared = %.4f)",
                        summary(full_model)$adj.r.squared),
      digits = 4)
```

**Addiction score is the one factor that stops being significant here**, and the reason is
the 0.91 correlation with gaming hours we flagged in the EDA. It is not adding a separate
effect; it is measuring gaming a second time. We report gaming hours as the real driver.

## ANOVA vs. correlation, side by side

```{r anova, eval = RUN_ANOVA}
# The slow block: aov() is fitting thousands of levels per model.
anova_list <- list(
  gaming     = aov(df$grades ~ factor(df$gaming_hours)),
  study      = aov(df$grades ~ factor(df$study_hours)),
  sleep      = aov(df$grades ~ factor(df$sleep_hours)),
  attendance = aov(df$grades ~ factor(df$attendance)),
  addiction  = aov(df$grades ~ factor(df$addiction_score))
)

anova_tables <- map_dfr(names(anova_list), function(n) {
  tab <- as.data.frame(summary(anova_list[[n]])[[1]])
  tab$term  <- rownames(tab)
  tab$model <- n
  tab
})
write_csv(anova_tables, "output/anova_results.csv")

comparison <- tibble(
  variable     = vars,
  obs_per_cell = cell_sizes$obs_per_cell,
  anova_p      = map_dbl(anova_list, ~ summary(.x)[[1]][["Pr(>F)"]][1]),
  correct_p    = cor_results$p_value
) %>%
  mutate(
    anova_says   = ifelse(anova_p   < 0.05, "Reject H0", "Fail to reject"),
    correct_says = ifelse(correct_p < 0.05, "Reject H0", "Fail to reject"),
    agrees       = ifelse(anova_says == correct_says, "", "DISAGREES")
  )

write_csv(comparison, "output/method_comparison.csv")

comparison %>%
  mutate(across(c(anova_p, correct_p), ~ format.pval(.x, digits = 3, eps = 1e-300))) %>%
  tbl(caption = "Where the wrong test changed the answer", digits = 2)
```

```{r anova-skipped, eval = !RUN_ANOVA, echo = FALSE, results = "asis"}
cat("*ANOVA comparison skipped — set `RUN_ANOVA <- TRUE` in the setup chunk to include it.*")
```

## Section conclusion

- **Study hours, gaming hours, sleep hours, and attendance** are statistically significant
  predictors of grades, both individually and when tested together.
- **Addiction score** is significant on its own but not once the other factors are
  included. It is a proxy for gaming hours, not a separate effect.
- **Attendance is significant** — the opposite of what our original ANOVA concluded, and
  the reason we changed method.

---

# Threshold analysis

*Source script: [`scripts/threshold_analysis.R`](../scripts/threshold_analysis.R)*

Objective 2 is to find the values *within* each factor where performance shifts. We use
two independent approaches so the cutoffs can be cross-checked: a decision tree, which
picks its own split points, and binned averages, which show where the mean grade steps
down.

## Letter grades

Converting to a standard letter scale lets a threshold be stated in language a reader
actually feels — "drops from A- to B+" rather than "drops 2.8 points".

```{r letters}
grade_breaks <- c(-Inf, 60, 63, 67, 70, 73, 77, 80, 83, 87, 90, 93, Inf)
grade_labels <- c("F", "D-", "D", "D+", "C-", "C", "C+", "B-", "B", "B+", "A-", "A")

to_letter <- function(x) cut(x, breaks = grade_breaks, labels = grade_labels,
                             right = FALSE)

df <- df %>% mutate(letter_grade = to_letter(grades))
write_csv(df, "data/cleaned_data_with_grades.csv")

df %>%
  count(letter_grade) %>%
  mutate(pct = round(100 * n / sum(n), 1)) %>%
  tbl(caption = "Letter grade distribution", digits = 1)
```

## Approach 1 — decision tree

`reaction_time_ms` is deliberately excluded. It is a *consequence* of gaming, not a cause
of grades, so letting it split would steal credit from `gaming_hours` and make the cutoffs
harder to interpret.

```{r tree, fig.width = 11, fig.height = 7}
predictors <- c("gaming_hours", "study_hours", "sleep_hours", "attendance",
                "addiction_score", "device_usage", "social_activity", "age",
                "stress_level", "gaming_genre")

tree_data <- df %>%
  select(grades, all_of(predictors)) %>%
  mutate(across(where(is.character), as.factor))

set.seed(432)
tree <- rpart(grades ~ ., data = tree_data, method = "anova",
              control = rpart.control(cp = 0.005, maxdepth = 4, minbucket = 100))

png("output/decision_tree.png", width = 1600, height = 1000, res = 150)
rpart.plot(tree, type = 4, extra = 101, box.palette = "RdYlGn",
           main = "Where grades split: regression tree on student features")
invisible(dev.off())

rpart.plot(tree, type = 4, extra = 101, box.palette = "RdYlGn",
           main = "Where grades split: regression tree on student features")
```

```{r tree-splits}
# rpart stores splits per node in the order: primary, competitors, surrogates.
# Walk the frame to pick out only the primary split of each internal node.
frame    <- tree$frame
splits   <- as.data.frame(tree$splits)
internal <- which(frame$var != "<leaf>")

split_rows <- integer(0)
idx <- 1
for (i in internal) {
  split_rows <- c(split_rows, idx)
  idx <- idx + 1 + frame$ncompete[i] + frame$nsurrogate[i]
}

node_ids <- as.numeric(rownames(frame))[internal]

tree_splits <- tibble(
  depth       = floor(log2(node_ids)) + 1,
  # rpart makes rownames unique by appending .1, .2 ... strip that back off
  variable    = sub("\\.[0-9]+$", "", rownames(splits)[split_rows]),
  split_value = round(splits$index[split_rows], 2),
  n_at_node   = frame$n[internal],
  mean_grade  = round(frame$yval[internal], 1)
) %>% arrange(depth, desc(n_at_node))

write_csv(tree_splits, "output/tree_splits.csv")
tbl(tree_splits, caption = "Tree split points — the candidate cutoffs", digits = 2)
```

The tree splits on study hours first, then on gaming hours — and it picks a *different*
gaming cutoff in each branch. That difference is itself a finding: the point at which
gaming starts to cost depends on how much a student studies, which is the definition of an
interaction, and it is why the modeling section tests one.

## Approach 2 — binned averages

Mean grade per bin with a 95% confidence interval, so a visible step down can be checked
against sampling noise before it gets called a threshold.

```{r bins, fig.height = 4.5}
bin_specs <- list(
  gaming_hours    = seq(0, 8, by = 1),
  study_hours     = seq(1, 10, by = 1),
  sleep_hours     = seq(4, 9, by = 1),
  attendance      = seq(60, 100, by = 5),
  addiction_score = seq(-5, 25, by = 5)
)

binned_summary <- function(var, breaks) {
  df %>%
    mutate(bin = cut(.data[[var]], breaks = breaks, include.lowest = TRUE)) %>%
    filter(!is.na(bin)) %>%
    group_by(bin) %>%
    summarise(n = n(), mean_grade = mean(grades),
              se = sd(grades) / sqrt(n()), .groups = "drop") %>%
    mutate(
      ci_low   = mean_grade - 1.96 * se,
      ci_high  = mean_grade + 1.96 * se,
      letter   = to_letter(mean_grade),
      variable = var,
      # Keep the within-variable ordering as a number. Stacking these tables would
      # otherwise merge the factor levels across variables and scramble which bin
      # follows which.
      bin_index = as.integer(bin),
      bin = as.character(bin)
    )
}

all_bins <- imap_dfr(bin_specs, ~ binned_summary(.y, .x))
write_csv(all_bins, "output/binned_means.csv")

for (v in names(bin_specs)) {
  d <- all_bins %>% filter(variable == v) %>%
    mutate(bin = fct_reorder(bin, bin_index))

  p <- ggplot(d, aes(x = bin, y = mean_grade)) +
    geom_col(fill = "steelblue", alpha = 0.85) +
    geom_errorbar(aes(ymin = ci_low, ymax = ci_high), width = 0.2) +
    geom_text(aes(label = letter), vjust = -0.8, size = 3.5) +
    labs(title = paste("Mean grade by", v),
         subtitle = "Error bars are 95% CIs; label is the letter grade of the bin mean",
         x = v, y = "Mean grade") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

  ggsave(paste0("output/binned_", v, ".png"), plot = p,
         width = 8, height = 5, dpi = 150)
  print(p)
}
```

## The cutoffs

The steepest drop between neighbouring bins is the plain-language threshold for each
factor.

```{r threshold-summary}
threshold_summary <- all_bins %>%
  group_by(variable) %>%
  arrange(bin_index, .by_group = TRUE) %>%
  mutate(drop = lag(mean_grade) - mean_grade,
         from_bin = lag(bin), from_letter = lag(letter)) %>%
  filter(!is.na(drop)) %>%
  slice_max(drop, n = 1) %>%
  ungroup() %>%
  transmute(
    variable,
    boundary   = paste(from_bin, "->", bin),
    grade_drop = round(drop, 1),
    shift      = paste(from_letter, "->", letter),
    n_after    = n,
    # A non-positive value means grades never actually step down for this factor,
    # so there is no threshold to report.
    threshold  = if_else(drop > 0, "yes", "none found")
  ) %>%
  arrange(desc(grade_drop))

write_csv(threshold_summary, "output/threshold_summary.csv")
tbl(threshold_summary, caption = "Steepest grade drop per factor", digits = 1)
```

## The whole distribution, not just the mean

```{r letter-mix, fig.width = 9, fig.height = 5}
p_mix <- df %>%
  mutate(bin = cut(gaming_hours, breaks = bin_specs$gaming_hours,
                   include.lowest = TRUE)) %>%
  filter(!is.na(bin)) %>%
  count(bin, letter_grade) %>%
  group_by(bin) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = bin, y = prop, fill = letter_grade)) +
  geom_col() +
  scale_fill_viridis_d(direction = -1) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Letter grade mix by daily gaming hours",
       x = "Gaming hours per day", y = "Share of students", fill = "Grade") +
  theme_minimal()

ggsave("output/letter_grade_by_gaming.png", p_mix, width = 9, height = 5, dpi = 150)
p_mix
```

A mean can move because everyone slipped a little or because a few students collapsed.
This chart settles it: the entire distribution slides down as gaming hours rise, so the
threshold is a shift in the whole population, not an artefact of a small tail.

---

# Predictive modeling

*Source script: [`scripts/modeling.R`](../scripts/modeling.R)*

Section 4 established which factors matter. This section asks the two questions a p-value
cannot answer: **what shape** is each relationship, and **how well does the model predict
students it has never seen**.

## Train / test split

```{r split}
set.seed(432)
train_idx <- sample(nrow(df), size = floor(0.7 * nrow(df)))
train <- df[train_idx, ]
test  <- df[-train_idx, ]

cat(sprintf("Training on %d students, testing on %d\n", nrow(train), nrow(test)))

# Grades were capped at 100 in cleaning, so a prediction above 100 cannot be right
# no matter what the formula says. Clamp before scoring.
rmse <- function(actual, predicted) sqrt(mean((actual - pmin(predicted, 100))^2))
r2   <- function(actual, predicted) {
  1 - sum((actual - pmin(predicted, 100))^2) / sum((actual - mean(actual))^2)
}
```

## Candidate models

Each model adds something the previous one lacked, so the comparison shows whether the
extra complexity earns its place.

```{r models}
models <- list(
  # Study hours alone, the strongest single predictor, as a floor to beat.
  study_only = grades ~ study_hours,

  # All five factors, straight line each. This is the model from section 4.3.
  main_effects = grades ~ gaming_hours + study_hours + sleep_hours + attendance +
                          addiction_score,

  # Gaming and addiction both bend rather than run straight (see the EDA scatterplots),
  # so add squared terms.
  with_curves = grades ~ gaming_hours + I(gaming_hours^2) + study_hours + sleep_hours +
                         attendance + addiction_score + I(addiction_score^2),

  # The tree split gaming at a different value in each branch, which is what an
  # interaction term represents: gaming costs more when study hours are low.
  with_interaction = grades ~ gaming_hours * study_hours + I(gaming_hours^2) +
                              sleep_hours + attendance + addiction_score +
                              I(addiction_score^2),

  # addiction_score correlates 0.91 with gaming_hours, so check what dropping it costs.
  no_addiction = grades ~ gaming_hours * study_hours + I(gaming_hours^2) +
                          sleep_hours + attendance
)

fits <- lapply(models, function(f) lm(f, data = train))

comparison_models <- imap_dfr(fits, function(m, name) {
  tibble(
    model      = name,
    n_terms    = length(coef(m)) - 1,
    train_rmse = round(rmse(train$grades, predict(m, train)), 3),
    test_rmse  = round(rmse(test$grades,  predict(m, test)),  3),
    test_r2    = round(r2(test$grades,    predict(m, test)),  4)
  )
})

write_csv(comparison_models, "output/model_comparison.csv")
tbl(comparison_models, caption = "Five candidate models, scored on the held-out test set",
    digits = 4)
```

`train_rmse` and `test_rmse` staying close is the sign we are **not overfitting**. If the
test error were much worse than the training error, the model would be fitting noise.

## Choosing the final model

Not simply the lowest `test_rmse`. Two of these land within 0.001 of each other, which is
noise from this particular split rather than a real difference, and picking on raw minimum
would hand us extra terms that do nothing. So we take the model with the **fewest terms
that gets within 1% of the best error**.

```{r choose}
best_rmse <- min(comparison_models$test_rmse)

best_name <- comparison_models %>%
  filter(test_rmse <= best_rmse * 1.01) %>%
  slice_min(n_terms, n = 1) %>%
  slice_min(test_rmse, n = 1) %>%
  pull(model)

cat(sprintf("Best test error: %.3f (%s)\n", best_rmse,
            comparison_models$model[which.min(comparison_models$test_rmse)]))
cat(sprintf("Chosen for parsimony, within 1%% of best: %s\n", best_name))
```

```{r final-model}
# Refit the chosen structure on all 8,000 students for the coefficients we report.
final_model <- lm(models[[best_name]], data = df)

coefs <- tidy(final_model) %>%
  mutate(across(where(is.numeric), ~ round(.x, 4)))

write_csv(coefs, "output/model_coefficients.csv")

coefs %>%
  mutate(p.value = format.pval(p.value, digits = 3, eps = 1e-300)) %>%
  tbl(caption = sprintf(
    "Final model coefficients, fit on all %d students (adjusted R-squared = %.4f, residual SE = %.2f grade points)",
    nrow(df), summary(final_model)$adj.r.squared, summary(final_model)$sigma),
    digits = 4)
```

## Diagnostics

Students sitting exactly on the 100-point cap are flagged separately throughout. Their
actual grade was truncated, so their residual is forced negative by the cap rather than by
the model being wrong; mixing them into the trend line would misrepresent the fit.

```{r diagnostics, fig.height = 5.5}
diag_df <- tibble(
  fitted   = fitted(final_model),
  residual = resid(final_model),
  censored = df$grades >= 100
)

p1 <- ggplot(diag_df, aes(x = fitted, y = residual)) +
  geom_point(aes(colour = censored), alpha = 0.25, size = 0.9) +
  geom_hline(yintercept = 0, colour = "red") +
  # Loess on the uncensored students only, so the cap does not drag the line down.
  geom_smooth(data = filter(diag_df, !censored), method = "loess",
              colour = "blue", se = FALSE) +
  scale_colour_manual(values = c("FALSE" = "grey30", "TRUE" = "darkorange"),
                      labels = c("FALSE" = "normal", "TRUE" = "at 100 point cap"),
                      name = NULL) +
  labs(title = "Residuals vs fitted values",
       subtitle = paste("Blue line is flat across the normal range, so the model fits.",
                        "\nThe orange diagonal is the grade cap, not model error."),
       x = "Predicted grade", y = "Residual") +
  theme_minimal() +
  theme(legend.position = "top")

ggsave("output/residuals_vs_fitted.png", p1, width = 8, height = 5.5, dpi = 150)
p1
```

```{r qq, fig.height = 5}
p2 <- ggplot(filter(diag_df, !censored), aes(sample = residual)) +
  stat_qq(alpha = 0.3) +
  stat_qq_line(colour = "red") +
  labs(title = "Normal Q-Q plot of residuals",
       subtitle = "Excludes students at the 100 point cap. Points on the line means the normality assumption holds",
       x = "Theoretical quantiles", y = "Sample quantiles") +
  theme_minimal()

ggsave("output/qq_residuals.png", p2, width = 8, height = 5, dpi = 150)
p2
```

```{r pred-actual, fig.width = 7, fig.height = 6}
test_pred <- tibble(
  actual    = test$grades,
  predicted = pmin(predict(final_model, test), 100)
)

p3 <- ggplot(test_pred, aes(x = actual, y = predicted)) +
  geom_point(alpha = 0.2) +
  geom_abline(slope = 1, intercept = 0, colour = "red") +
  labs(title = "Predicted vs actual grades, held-out test set",
       subtitle = "Red line is a perfect prediction",
       x = "Actual grade", y = "Predicted grade") +
  theme_minimal()

ggsave("output/predicted_vs_actual.png", p3, width = 7, height = 6, dpi = 150)
p3
```

---

# Conclusions

## What the analysis found

```{r findings-table, echo = FALSE}
tibble(
  Question = c(
    "Which factors relate to grades?",
    "Which factor matters most?",
    "What about addiction score?",
    "Where are the thresholds?",
    "Is the relationship linear?",
    "How accurate is the model?"
  ),
  Answer = c(
    "Study hours, gaming hours, sleep hours and attendance are all significant (p < 0.05), individually and jointly.",
    sprintf("Study hours — %.2f grade points per additional hour, the largest single effect.",
            cor_results$slope[cor_results$variable == "study_hours"]),
    sprintf("Significant alone, but not once the others are controlled for. It correlates %.2f with gaming hours, so it re-measures gaming rather than adding a separate effect.",
            cor(df$addiction_score, df$gaming_hours)),
    sprintf("The tree and the binned means agree. Steepest drop overall: %s at the %s boundary (%s grade points).",
            threshold_summary$variable[1], threshold_summary$boundary[1],
            threshold_summary$grade_drop[1]),
    "No — gaming and addiction both bend, so squared terms improve the fit, and the gaming penalty is worse for students who study less (a real interaction).",
    sprintf("Held-out test RMSE of %.2f grade points, test R-squared %.3f. Train and test error are close, so the model is not overfitting.",
            comparison_models$test_rmse[comparison_models$model == best_name],
            comparison_models$test_r2[comparison_models$model == best_name])
  )
) %>%
  tbl(caption = "Summary of findings")
```

## Limitations

```{r ceiling, echo = FALSE}
ceiling_n <- sum(df$grades >= 100)
```

1. **Censored outcome.** We capped grades at 100 in cleaning, so `r ceiling_n` students
   (`r sprintf("%.1f%%", 100 * ceiling_n / nrow(df))`) sit exactly on the ceiling. The
   model cannot distinguish a student who would have scored 101 from one who would have
   scored 118. A censored (tobit) regression handles this properly but is beyond the scope
   of this course project, so we flag it rather than paper over it.
2. **The data is almost certainly simulated.** An adjusted R-squared near
   `r sprintf("%.2f", summary(final_model)$adj.r.squared)` is not something real survey
   data produces, and a 0.91 correlation between two supposedly independent variables is
   far too clean. Read these results as a demonstration of the method working on the data
   as given, not as a real-world finding about students.
3. **Observational, not causal.** Students who game more also study less. The coefficients
   describe association. Nothing here proves that a student who cut their gaming would see
   their grade rise; that claim would need an experiment.
4. **Self-reported measures.** Gaming hours, study hours, sleep, and the addiction score
   are the kind of variables that are systematically misreported in surveys, in
   predictable directions.

## Reproducing this report

Everything above runs from the raw CSV. To regenerate the HTML and every file in
`output/`:

```{r reproduce, eval = FALSE}
rmarkdown::render("report/final_report.Rmd")
```

or, from a terminal at the project root:

```
Rscript scripts/render_report.R
```

```{r session, echo = FALSE}
tibble(
  Item = c("R version", "Platform", "Run date", "Rows analysed", "Key packages"),
  Value = c(
    R.version.string,
    R.version$platform,
    format(Sys.time(), "%Y-%m-%d %H:%M"),
    format(nrow(df), big.mark = ","),
    paste(c("tidyverse", "broom", "rpart", "rpart.plot", "corrplot"), collapse = ", ")
  )
) %>%
  tbl(caption = "Session information")
```
